---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# psacomp

<!-- badges: start -->
<!-- badges: end -->

*psacomp* is a package for computing *Principal Subsimplex Analysis* introduced in *Principal Subsimplex Analysis* (2025+), Hyeon Lee, Kassel Liam Hingee, Janice L. Scealy, Andrew T. A. Wood, Eric Grunsky, and J. S. Marron.

## Installation
The package `psacomp` can be installed by running the following code.

```{r, eval = F}
library(devtools)
devtools::install_github("haneone33/psacomp")
```

## PSA Basics

First we create an example compositional data set with 7 variables and 20 data points.
```{r, include = TRUE}
set.seed(1)
X = matrix(runif(20*7),20,7)
X = sweep(X, 1, rowSums(X), '/')
```

Main function of *psacomp* is `psa()` which takes `type` and `X` as arguments. Available options for `type` are `type = 's'` for PSA-S and `type = 'o'` for PSA-O.

```{r, incdlue = TRUE}
library(psacomp)
X.psas = psa(type = 's', X = X)
X.psao = psa(type = 'o', X = X)
```

## Output Details
`psa()` outputs a list of information calculated through PSA. The information is summarized in the following table. We assume `X` is a data matrix with $n$ rows and $D$ columns.

```{r table1, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
| Name       | Type         | Dimension   | Description                        |
|--------|:------------:|:-------:|------------------------------------|
| Vhat   | list         | $D$         | `vertices$'r=i'` is an $i\\times D$ matrix of vertices of representation with $i$ vertices, each row representing a vertex   |
| Xhat | list         | $D$         | `pts.approx$'r=i'` is an $n\times i$ matrix of rank $i-1$ representation, in the new vertex system of `vertices$'r=i'`        |
| Xhat_reduced | list         | $D$         | `pts$'r=i'` is an $n\\times D$ matrix of rank $i-1$ representation, in the original vertex system of $D$ unit vertices       |
| X      | matrix | $n\\times D$ | given data matrix |
| residuals | list | $D-1$ | residuals |
| scores     | matrix       | $n\\times (D-1)$ | scores |
| RSS  | vector       | $D-1$         | residual sum of squares            |
| backwards_mean | vector | $D$ | the backwards mean |
| loadings   | matrix       | $D\\times (D-1)$ | a matrix of loading vectors, each column representing a loading vector                                                  |
| construction_info | data frame   | $D\\times 3$ | construction information. contains index of merged vertices and merge weights                                              |
"
cat(tabl)
```

